```{r echo = FALSE}
# This block needs cache=FALSE to set fig.width and fig.height, and have those
# persist across cached builds.
source("utils.R", local = TRUE)
```

# R Programming language primer

The R programming language comprises 6 major types of [R objects](https://rstudio-education.github.io/hopr/r-objects.html)

  - [Vectors](https://rstudio-education.github.io/hopr/r-objects.html#atomic-vectors)
  - [Matrix](https://rstudio-education.github.io/hopr/r-objects.html#matrices)
  - [Data Frame](https://rstudio-education.github.io/hopr/r-objects.html#data-frames)
  - [Lists](https://rstudio-education.github.io/hopr/r-objects.html#lists)
  - [Array](https://rstudio-education.github.io/hopr/r-objects.html#arrays)
  - [Factors](https://rstudio-education.github.io/hopr/r-objects.html#factors)


## Vectors

The vector type is the fundamental data type in R. The elements of a vector must all have the same mode (data type) of the following type:

  - _Scalars_ are individual numbers, they actually are one-element vectors e.g. `x <- 10`
  - _Character_ strings are single-element vectors of mode ”character” e.g. `y <- "hello"`


```{r}
x <- c(1, 2, 3, 4, 5, 6)
x
x <- 1:6
x
x[1]
text <- c("Hello",  "World")
text
names(x) <- paste("name", 1:6, sep="")
x
```

### Vectors in action

```{r}
z_scores <- seq(-3, 3, by = .1)
z_scores
norm.dist <- dnorm(z_scores)
norm.dist
mean(norm.dist)
median(norm.dist)
sd(norm.dist)
```


## Matrix

A matrix is a rectangular array of numbers (technically a vector with a number of rows and a number of columns)

```{r}
m <- matrix(x, nrow = 2)
m

m <- matrix(x, nrow = 2, byrow = TRUE)
m
```

### Matrices in action

```{r}
normMat <- matrix(rnorm(10),nrow=5,ncol=20)
normMat[1:5,1:5]
rownames(normMat) <- paste("Gene", 1:5, sep="_")
normMat[1:5,1:5]
#Calculate mean for each matrix row
apply(normMat, MARGIN=1, mean)
#Calculate mean for each matrix column
apply(normMat, MARGIN=2, mean)
```


## Data Frame

A data frame is a data set that contains data of different modes (i.e. different types). A data frame is a list with each component of a list being a vector corresponding to a column in the data.

```{r}
df <- data.frame(cancerType = c("PDAC", "PDAC", "PDAC"),  
  KRAS_status = c("wt", "mut", "mut"), Subtype = c("Classical", "Classical", "Subtype"))
df
```


### Data Frames in action

```{r}
#Extract cancerType
df[,1]
#Extract cancerType
df$cancerType
#Assigning new data
df$Sex  <- c("Male", "Female", "Male")
df
#Excluding factors
df <- data.frame(cancerType = c("PDAC", "PDAC", "PDAC"),  
  KRAS_status = c("wt", "mut", "mut"), Subtype = c("Classical", "Classical", "Subtype"), stringsAsFactors = FALSE)
df$cancerType
```


## Factors

Factors are used to represent categorical data. Factors can be ordered or unordered and are an important class for statistical analysis and for plotting.
```{r}
subtype <- factor(c("Classical", "Basal", "Basal", "Classical"))
subtype
levels(subtype)
table(subtype)
```


### Factors in action

```{r}
colours <- c("red", "green")
colours[subtype]
```


## List

Lists can contain items comprising different data types and are similar to a C struct in C++.
```{r}
list1 <- list(1:20, "KRAS", list(1:6, 7:10))
list1
```


### Lists in action

```{r}
list2 <- list(rnorm(10), rnorm(10), rnorm(10))
#Access first entry of list2
list2[1]
#Access vector of values in first entry of list2
list2[[1]]
#Finding mean for each vector series in list2 - returns a list of means
lapply(list2, mean)
#Finding mean for each vector series in list2 - returns a vector of means
sapply(list2, mean)
```


## Array

Arrays are matrices with more than 2 dimensions.

```{r}
ar <- array(c(11:14, 21:24, 31:34), dim = c(2, 2, 3))
ar
```


## Creating functions


```{r}
#Function to select the top n most variable genes
rnaSelectTopVarGenes <- function(normCounts, n=2000){
  mad <- apply(normCounts, 1, mad)
  ords <- names(sort(mad, decreasing=TRUE))
  select <- normCounts[ords[1:n],]
  select
}

top2varGenes <- rnaSelectTopVarGenes(normMat, n=2)
head(top2varGenes)
```



## Loading Data

Loading data from a comma separated file can be easily achieved using the `read.delim` or  `read.csv` functions. There are many alternatives!

```{r}
moffitt <- read.delim("course/Moffitt_gene_signature.csv", 
                      sep=",", stringsAsFactors = FALSE)
head(moffitt)
class(moffitt)
moffitt$symbol
```

## Saving Data

Saving data is also simple. You can write data to comma separated (csv) files using the `write.table` function.

The best way to store objects from R is within `.RData` files. `.RData` files are specific to R and can store as many objects as required.

```{r eval=FALSE}
#Write data to a tab separate file called Moffitt_gene_signature.tsv
write.table(moffitt, "course/Moffitt_gene_signature.tsv", 
            sep="\t", quote = FALSE, row.names = FALSE)
#Save moffitt to a RData file called moffittSig.RData
save(moffitt,x,y,z,file="course/moffittSig.RData")
#Load themoffitt data.frame into the current environment
load("course/moffittSig.RData")
```

## Saving plots

How do we save complex plots like the one below?

```{r, fig.cap="Acceptance and rejection regions in a t-test distribution.", fig.width=6, fig.height=4}
df <- 6 - 1
f <- function(x) {
  dt(x, df)
}
plot(f, -4, 4, xlab = "x-axis", ylab = "T Density dt(x)")
ci <- c(qt(0.025, df), qt(0.975, df))
x <- seq(-4, ci[1], 0.01)
y <- seq(ci[2], 4, 0.01)
polygon(c(ci[1], x, ci[1]), c(0, 0, f(x)), col = "red")
polygon(c(ci[2], y, ci[2]), c(f(y), 0, 0), col = "red")
arrows(-3, 0.25, -3, 0.05)
text(-3, 0.3, "Rejection Region")
arrows(3, 0.25, 3, 0.05)
text(3, 0.3, "Rejection Region")
arrows(0, 0.15, 0, 0.05)
text(0, 0.2, "Acceptance Region")
```


It is easy to save the above plot using the `pdf()` and `dev.off()` functions as follows. All of the plotting commands contained between the `pdf() & dev.off()` functions will be sent to file. It is also possible to use the `png()` function if you would like to create png's instead! Check out the `grDevices` package for other alternatives.

```{r eval=FALSE}
pdf("course/normalDist.pdf", height=6, width=4)
  df <- 6 - 1
  f <- function(x) {
    dt(x, df)
  }
  plot(f, -4, 4, xlab = "x-axis", ylab = "T Density dt(x)")
  ci <- c(qt(0.025, df), qt(0.975, df))
  x <- seq(-4, ci[1], 0.01)
  y <- seq(ci[2], 4, 0.01)
  polygon(c(ci[1], x, ci[1]), c(0, 0, f(x)), col = "red")
  polygon(c(ci[2], y, ci[2]), c(f(y), 0, 0), col = "red")
  arrows(-3, 0.25, -3, 0.05)
  text(-3, 0.3, "Rejection Region")
  arrows(3, 0.25, 3, 0.05)
  text(3, 0.3, "Rejection Region")
  qrrows(0, 0.15, 0, 0.05)
  text(0, 0.2, "Acceptance Region")
dev.off()
```

